---
title: 'Spatiotemporal Stats'
author: "Savita Upadhyay"
subtitle: Spatiotemporal Stats Lab 
output:
  pdf_document: default
---

```{r setup, echo=T, message=F, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache = TRUE)
library(STRbook)
library(ggplot2)
library(dplyr)
library(tidyr)
library(sp)
library(spacetime)
library(lubridate)
```

## Questions

For the SST data loaded using the code below:

1. generate a data frame with the Empirical Spatial Means per decade (1970-1979, 1980-1989, 1990-2002) and plot them with one panel per decade

2. generate a spatial plot for the **yearly**  SST 95th quantile for the years 1980, 1990, 2000 having one panel per year

3. Obtain a Hovmoller plot for these data

4. Calculate the EOFs for the *SST* dataset.  How many EOFs would you retain?


\scriptsize
```{r}
data("SSTlandmask", package = "STRbook")
data("SSTlonlat", package = "STRbook")
data("SSTdata", package = "STRbook")


#combining the data frame
#SSTdata_lonlat <- cbind(SSTdata,
#                          SSTlonlat)

#remove years that are not complete
rm_rows <- which(SSTlandmask == 1)
SSTdata <- SSTdata[-rm_rows, 1:396]
SSTlonlat <- SSTlonlat[-rm_rows, 1:2]


SSTdata_n <- cbind(SSTlonlat, SSTdata)
SST_df <- gather(SSTdata_n, date, sst, -lon, -lat) 

date_grid <- expand.grid(Month = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug","Sep", "Oct", "Nov", "Dec"),
                         Year = 1970:2002,
stringsAsFactors = FALSE) 

date_grid$date <- paste0("V", 1:396)
date_grid$t <- seq(1,396,1)

SST_df <- left_join(SST_df, date_grid)%>% 
  mutate(decade=cut(Year, breaks=c(1969,1979,1989,1999,2009),
                    dig.lab=4,right = F,include.lowest = F)) 
SST_df$date<-NULL
colnames(SST_df)<-c("lon","lat","values","Month","Year","decade","t")
```

1. generate a data frame with the Empirical Spatial Means per decade  and plot them with one panel per decade

\scriptsize
```{r echo=T, eval=T, message=F, fig.height=2.5, fig.width=3, fig.align='center'}

summ1 <- SST_df %>% 
  group_by(lon,lat,Year) %>% 
  summarise(proc_means=mean(values))
summ1


summ2 <- SST_df %>% 
  group_by(lon,lat,decade) %>% 
  summarise(proc_means=mean(values))
summ2

SST_df %>% 
  group_by(lon,lat,decade) %>%
  summarise(z=mean(values)) %>% 
  ggplot() + 
  geom_point(aes(x=lon,y=lat,colour=z),size=2) +
  col_scale(name = "degF") + 
  xlab("Longitude (deg)") + 
  ylab("Latitude (deg)") 

SST_df %>% 
  group_by(lon,lat,Year) %>%
  summarise(z=mean(values)) %>% 
  ggplot() + 
  geom_point(aes(x=lon,y=lat,colour=z),size=2) +
  col_scale(name = "degF") + 
  xlab("Longitude (deg)") + 
  ylab("Latitude (deg)") 

```
2. generate a spatial plot for the **yearly**  SST 95th quantile for the years 1980, 1990, 2000 having one panel per year

\scriptsize
```{r echo=T, eval=T, message=F, fig.height=2.5, fig.width=3, fig.align='center'}


summ_quantile <- SST_df %>%
   filter(Year %in% c(1980,1990,2000))%>%
   group_by(lon,lat,Year) %>% 
  summarise(proc_quantile=quantile(values,0.95))
summ_quantile 

SST_df_1 <- subset(SST_df, Year %in% c(1980,1990,2000))

SST_df_1 %>% 
  group_by(lon,lat,Year) %>%
  summarise(z = quantile(values,0.95)) %>% 
  ggplot() + 
  geom_point(aes(x=lon,y=lat,colour=z),size=2) +
  col_scale(name = "degF") + 
  xlab("Longitude (deg)") + 
  ylab("Latitude (deg)") 

seq_along = c(1980,1990,2000)

plot_quantile_yearly <- function(i,SST_df_1) {  
    # for each longitudinal strip 
SST_df_1 %>% 
   filter(Year ==seq_along[i])%>% 
  group_by(lon,lat) %>%
  summarise(z = quantile(values,0.95)) %>% 
  ggplot() + 
  geom_point(aes(x=lon,y=lat,colour=z),size=2) +
  col_scale(name = "degF") + 
  xlab("Longitude (deg)") + 
  ylab("Latitude (deg)")     
}
par(mfrow=c(1,3))
plot_quantile_yearly(1,SST_df_1)
plot_quantile_yearly(2,SST_df_1)
plot_quantile_yearly(3,SST_df_1)
```

3. Obtain a Hovmoller plot for these data

```{r}
lim_lat <- range(as.numeric(SST_df$lat))
lim_t <- range(as.numeric(SST_df$Year))
lim_t1 <- range(as.numeric(SST_df$decade))
lat_axis <- seq(lim_lat[1],
lim_lat[2],
length=25) 
t_axis <- seq(lim_t[1], lim_t[2],
length=100)
lat_t_grid <- expand.grid(lat = lat_axis,
Year= t_axis)

SST_df_grid <- SST_df
dists <- abs(outer(SST_df$lat, lat_axis, "-"))
SST_df$lat <- lat_axis[apply(dists, 1, which.min)]

SST_df_lat_Hov <- group_by(SST_df_grid, lat, Year) %>% summarise(z = mean(values))

Hovmoller_lat <- ggplot(SST_df_lat_Hov) + 
geom_tile(aes(x = lat, y = Year, fill = z)) + 
fill_scale(name = "degC") + scale_y_reverse() + ylab("Years") + xlab("Latitude (degrees)") + theme_bw()

Hovmoller_lat 
```
4. Calculate the EOFs for the *SST* dataset. How many EOFs would you retain?

```{r fig.width=5, fig.height=3, fig.align="center"}
## Put data into space-wide form
nyr=(2002-1970)
colnames(SSTdata) <- paste(1, rep(1:12,nyr), 
                           rep(1970:2002,each=12), sep="-")

# append coordinates
SSTdata <- cbind(SSTlonlat[-rm_rows,],SSTdata[-rm_rows,])
Z <- t(SSTdata[,-(1:2)])
Cz <- cov(Z)
Spec <- eigen(Cz)

# eigenvalues
lambda <- Spec$values
# eigenvectors
Psi <- Spec$vectors #EOF's

# variance explained
propvar <- lambda/sum(lambda)
tibble(index=1:length(lambda),cumprop=propvar) %>% 
  filter(index<21) %>% 
  ggplot() +
  geom_line(aes(x=index,y=cumprop)) +
  geom_point(aes(x=index,y=cumprop),size=2.5) +
  ggthemes::theme_economist() +
  ylab(expression(lambda[k]/sum(lambda[j],j,))) +
  labs(title="Proportion of variance explained")
```


```{r fig.height=2,fig.width=5, fig.align="center"}
# Plot first 4 EOF's

EOF4 <- Psi[,1:4]
colnames(EOF4) <- paste0("psi",1:4)
EOF4 <- bind_cols(SSTdata[,1:2],as_tibble(EOF4[,1:4])) %>% 
  pivot_longer(cols=starts_with("Psi"),
               names_to="EOF",
               values_to="value")

ggplot(subset(EOF4,EOF=="psi1"),aes(x=lon,y=lat))+
  geom_tile(aes(fill=-value)) +
  fill_scale(name = "psi1") +
  theme_bw()
  
```

```{r fig.height=2,fig.width=5, fig.align="center"}
# Plot first 4 EOF's
ggplot(subset(EOF4,EOF=="psi2"),aes(x=lon,y=lat))+
  geom_tile(aes(fill=-value)) +
  fill_scale(name = "psi2") +
  theme_bw()
  
```

```{r fig.height=2,fig.width=5, fig.align="center"}
# Plot first 4 EOF's
ggplot(subset(EOF4,EOF=="psi3"),aes(x=lon,y=lat))+
  geom_tile(aes(fill=-value)) +
  fill_scale(name = "psi3") +
  theme_bw()
```

```{r fig.height=2,fig.width=5, fig.align="center"}
# Plot first 4 EOF's
ggplot(subset(EOF4,EOF=="psi4"),aes(x=lon,y=lat))+
  geom_tile(aes(fill=-value)) +
  fill_scale(name = "psi4") +
  theme_bw()
```

```{r}
# PC time series (one per column)
A <- (Z%*%Psi)%*%diag(1/sqrt(round(lambda,8)))
colnames(A) <- paste0("a",1:ncol(A),"(t)")
A4 <- as_tibble(A[,1:4]) %>% 
  mutate(date=dmy(rownames(A))) %>% 
  pivot_longer(cols=starts_with("a"),
               names_to="pcts",
               values_to="value")
  
ggplot(A4) +
  geom_line(aes(x=date,y=(-1)*value)) +
  facet_wrap(~pcts) +
  ggthemes::theme_economist()
```



